Discrimination in police award nomination in Chicago

### 2024-02-14 outputs/paper/paper.qmd
Deleted improper files

---
title: "Discrimination in police award nomination in Chicago"
author: 
  - Hyungsoo Park
  - Jeongwoo Kim
thanks: "https://github.com/kakaomonk/discrimination_in_police_award_nominations_chicago"
date: today
date-format: long
abstract: "This study investigates the racial and gender disparities in award nomination in the police department in Chicago by drawing a dataset that includes policing activities and award nominations. The original study highlighted both economically and statistically significant gaps between black and white officers, as well as male and female, in terms of award nominations. This paper aims to verify the existing biases within the policing organization and culture and explore the implications. Our study not only validates the original findings but also delves into the disparities in years of experience, and other potential variables ---. This replication contributes to the broader discourse on persistent disparities within policing culture by providing a thorough examination of the potential factors affecting award nominations."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

By using a dataset comprised of policing activity, award nominations, and officer’s background, this research bridges the gap between labor market discrimination, the analysis focuses on departmental award nominations as a key metric of internal recognition to examine a potential bias against minority groups - black and female officers. We use award nomination rather than promotion as a performance indicator since promotion is rare in policing departments; therefore, this approach offers a more frequent and discernible measure of how the department values each officer. This study’s innovations include a detailed regression analysis from the original paper and discussing other potential variables which were not discussed in the original paper. This replication study aims to reaffirm the findings of Rim Et al (2020) by replicating the methodological framework used in her paper to examine the results of award nominations within the CPD, controlling policing activities such as uses of force, number of civilian complaints, and number of arrests. By doing this, it seeks to disclose the internal police department dynamics, specifically how racial and gender disparities exist in law enforcement culture. This replication contributes to the broader discourse on labor market discrimination, particularly within the context of law enforcement, by providing a rigorous examination of the factors influencing award nominations and the potential impact of these disparities on minority representation and policing outcomes.

We find a sizable and statistically significant gap in award nominations between black and white officers as well as female and male officers, confirming our initial hypothesis that there are racial and gender disparities within policing departments. This indicates a persistent inequity with black officers receiving approximately 28 percent fewer nominations than their white counterparts, and female officers receiving 34 percent fewer nominations than their male counterparts. These disparities carry implications for the existence of discrimination within the policing department. The intersectionality of race and gender further makes this a serious issue as black female officers face a compounded disadvantage. This research has its implications not only in academia but also in real life. The findings inform policy debates and interventions to make the police force more equitable and effective. This paper is structured to first introducing a detailed analysis of the CPD dataset, followed by presenting our methodological approach and regression model findings and results. We then introduce our discussion in terms of relationships between complaints and award nominations and arrests and award nominations separately. Our work contributes to the disclose of labor market discrimination within law enforcement, urging for a more equitable recognition system. 
# Data

@@

```{r}
#| label: Table 1
#| fig-cap: Table
#| echo: false
month_data <- read_dta(here("inputs/data/officer_month_data.dta"))
year_data <- read_dta(here("inputs/data/officer_year_data.dta"))

@@

where yit represents the number of award received by an officer i in year t. ‘Black’, ‘Female’, ‘Hispanic’, ‘Asian’ and ‘NatAm’ are binary indicator variables equal to one that indicate whether the officer belongs to one of these racial or gender groups respectively, with white male officers are a reference group. Xit includes control variables like cohort, year, birth year fixed effects along with tenure, complaints, arrests, and use of force. the authors from the original paper state that they have included squared and cubed terms for tenure, complaints, arrests and use of force to consider nonlinear effects. Standard errors are also grouped by cohort for accuracy. However, in this paper, we have the difficulty in reproducing exactly the same with the original paper in these regards because information is not published in terms of errors and second-order and third-order terms. We expect the coefficients of the variables of our findings to be slightly different to the original paper.

@@

```{r}
#| echo: false
bl_fm <- c("Black", "Female", "Black x Female", "Mean")
first <- c(-10.368, -12.826, "", 35.307)
second <- c(-10.180, -14.879, 8.315, 37.583)
final_model <- data.frame(bl_fm, first, second)
kable(final_model, col.names = c("", "(1)", "(2)"))
```

In the third column in table 2, we introduce an important variable considering race and gender by interacting black and female. The coefficient for female increase slightly for female officer to -2 nominations along with an interaction effect valued at 8.3. the results suggest that being a black officer, regardless of gender, correlates with 28 percent reduction in award nominations to their non-black male counterparts. On the other hand, our analysis highlights the importance of gender among non-black officers as non-black female officers receiving 24 percent fewer nominations when compared to non-black male officers. This intersectionality of race and gender demands more attention. Black female officers face a compounded disadvantage, indicative of systemic issue that raises awareness and needs a change in the culture and policy in the departmental culture.

@@

The original paper does not expend its investigation by concluding that gender discrepancy exists among non-black officers. From the dataset, we have put an emphasis on the variable: complaints and wondered if there is any correlation between the award nominations for an officer and the number of complaints received and see if this affects the disparity between the majority-minority groups. We hypothesize that there is a negative correlation between the two variables; the more complaints an officer receive, the less nominations are awarded to the officer. If this pattern is found and statistically significant, this implies that the police department’s reward system is responsive to complaints, which may discourage negative behavior among officers. We have designed a scatterplot to represent the relationship between the two variables.

```{r}
#| label: Figure 1 ()
#| fig-cap: Bills of penguins
#| echo: false

comp_awd <- ggplot(all_officer, aes(x = Complaints, y = awd_perf)) + 
  ggtitle("Plot of Award Nomination by Number of Complaints") +
  xlab("Number of Complaints") + ylab("Award Nomination") +
  geom_point(shape=18, color="blue")

@@

Figure 1 correlates these two variables with Complaints in the x-xis and award nominations on the y-axis. The plot shows distinct vertical clusters of points at each value of the number of complaints, a discrete variable. We find that there is a lack of a clear trend that indicates a strong negative correlation across the entire range of the data. Instead, the dense concentration of points at the lower numbers of complaints (0 to 2) suggests that officers with fewer complaints receive a wider range of award nominations from 0 to 30 in extreme cases. However, this range seems to decrease as the number of complaints increases, with fewer award nominations at the higher numbers of complaints (4 and 5). This may provide some support for the hypothesis that more complaints lead to fewer award nominations, particularly at the higher end of the complaint scale. At lower complaint scale, the likelihood of receiving award nominations may vary depending on each officer, but as it gets to the higher complaint scale, the likelihood of receiving many award nominations seem to decrease. However, it is also important to note that there may be a pattern where officers with many complaints receive fewer number of awards, there is no clear evidence of a linearity or proportionality.

@@

```{r}
#| label: Figure 2 ()
#| fig-cap: Bills of penguins
#| echo: false

arr_awd <- ggplot(all_officer, aes(x = arrest, y = awd_perf)) + 
  ggtitle("Plot of Award Nomination by Number of Arrests") +
  xlab("Number of Arrests") + ylab("Award Nomination") +
  geom_point(shape=18, color="red")

>>>>

---
title: "Discrimination in police award nomination in Chicago"
author: 
  - Hyungsoo Park
  - Jeongwoo Kim
thanks: "https://github.com/kakaomonk/discrimination_in_police_award_nominations_chicago"
date: today
date-format: long
abstract: "This study investigates the racial and gender disparities in award nomination in the police department in Chicago by drawing a dataset that includes policing activities and award nominations. The original study highlighted both economically and statistically significant gaps between black and white officers, as well as male and female, in terms of award nominations. This paper aims to verify the existing biases within the policing organization and culture and explore the implications. Our study not only validates the original findings but also delves into the disparities in years of experience, and other potential variables ---. This replication contributes to the broader discourse on persistent disparities within policing culture by providing a thorough examination of the potential factors affecting award nominations."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

By using a dataset comprised of policing activity, award nominations, and officer’s background, this research bridges the gap between labor market discrimination, the analysis focuses on departmental award nominations as a key metric of internal recognition to examine a potential bias against minority groups - black and female officers. We use award nomination rather than promotion as a performance indicator since promotion is rare in policing departments; therefore, this approach offers a more frequent and discernible measure of how the department values each officer. This study’s innovations include a detailed regression analysis from the original paper and discussing other potential variables which were not discussed in the original paper. This replication study aims to reaffirm the findings of Rim Et al (2020)----------- by replicating the methodological framework used in her paper to examine the results of award nominations within the CPD, controlling policing activities such as uses of force, number of civilian complaints, and number of arrests. By doing this, it seeks to disclose the internal police department dynamics, specifically how racial and gender disparities exist in law enforcement culture. This replication contributes to the broader discourse on labor market discrimination, particularly within the context of law enforcement, by providing a rigorous examination of the factors influencing award nominations and the potential impact of these disparities on minority representation and policing outcomes.

We find a sizable and statistically significant gap in award nominations between black and white officers as well as female and male officers, confirming our initial hypothesis that there are racial and gender disparities within policing departments. This indicates a persistent inequity with black officers receiving approximately 28 percent fewer nominations than their white counterparts, and female officers receiving $34$ percent fewer nominations than their male counterparts. These disparities carry implications for the existence of discrimination within the policing department. The intersectionality of race and gender further makes this a serious issue as black female officers face a compounded disadvantage. This research has its implications not only in academia but also in real life. The findings inform policy debates and interventions to make the police force more equitable and effective. This paper is structured to first introducing a detailed analysis of the CPD dataset, followed by presenting our methodological approach and regression model findings and results. We then introduce our discussion in terms of relationships between complaints and award nominations and arrests and award nominations separately. Our work contributes to the disclose of labor market discrimination within law enforcement, urging for a more equitable recognition system.

# Data

@@

```{r}
#| label: all-officers-characteristics
#| echo: false
#| tbl-cap: "All officers' characteristics"

month_data <- read_dta(here("inputs/data/officer_month_data.dta"))
year_data <- read_dta(here("inputs/data/officer_year_data.dta"))

@@

where $y_{it}$ represents the number of award received by an officer $i$ in year $t$. ‘Black’, ‘Female’, ‘Hispanic’, ‘Asian’ and ‘NatAm’ are binary indicator variables equal to one that indicate whether the officer belongs to one of these racial or gender groups respectively, with white male officers are a reference group. Xit includes control variables like cohort, year, birth year fixed effects along with tenure, complaints, arrests, and use of force. the authors from the original paper state that they have included squared and cubed terms for tenure, complaints, arrests and use of force to consider nonlinear effects. Standard errors are also grouped by cohort for accuracy. However, in this paper, we have the difficulty in reproducing exactly the same with the original paper in these regards because information is not published in terms of errors and second-order and third-order terms. We expect the coefficients of the variables of our findings to be slightly different to the original paper.

@@

```{r}
#| echo: false
#| tbl-cap: "Racial and Sex Dispaties in Annual Nominations"
bl_fm <- c("Black", "Female", "Black x Female", "Mean")
first <- c(-10.368, -12.826, "", 35.307)
second <- c(-10.180, -14.879, 8.315, 37.583)
final_model <- data.frame(bl_fm, first, second)
kable(final_model, col.names = c("", "(1)", "(2)"))
```

In the third column in table 2, we introduce an important variable considering race and gender by interacting black and female. The coefficient for female increases slightly for female officer to -2 nominations along with an interaction effect valued at 8.3. The results suggest that being a black officer, regardless of gender, correlates with 28 percent reduction in award nominations to their non-black male counterparts. On the other hand, our analysis highlights the importance of gender among non-black officers as non-black female officers receiving 24 percent fewer nominations when compared to non-black male officers. This intersectionality of race and gender demands more attention. Black female officers face a compounded disadvantage, indicative of systemic issue that raises awareness and needs a change in the culture and policy in the departmental culture.

@@

The original paper does not extend its investigation by concluding that gender discrepancy exists among non-black officers. From the dataset, we have put an emphasis on the variable: complaints and wondered if there is any correlation between the award nominations for an officer and the number of complaints received and see if this affects the disparity between the majority-minority groups. We hypothesize that there is a negative correlation between the two variables; the more complaints an officer receive, the less nominations are awarded to the officer. If this pattern is found and statistically significant, this implies that the police department’s reward system is responsive to complaints, which may discourage negative behavior among officers. We have designed a scatterplot to represent the relationship between the two variables.

```{r}
#| label: award-nomination-by-number-of-complaints
#| fig-cap: Award Nomination by Number of Complaints
#| echo: false

comp_awd <- ggplot(all_officer, aes(x = Complaints, y = awd_perf)) + 
  xlab("Number of Complaints") + ylab("Award Nomination") +
  geom_point(shape=18, color="blue")

@@

Figure 1 correlates these two variables with Complaints in the x-axis and Award Nominations on the y-axis. The plot shows distinct vertical clusters of points at each value of the number of complaints, a discrete variable. We find that there is a lack of a clear trend that indicates a strong negative correlation across the entire range of the data. Instead, the dense concentration of points at the lower numbers of complaints (0 to 2) suggests that officers with fewer complaints receive a wider range of award nominations from 0 to 30 in extreme cases. However, this range seems to decrease as the number of complaints increases, with fewer award nominations at the higher numbers of complaints (4 and 5). This may provide some support for the hypothesis that more complaints lead to fewer award nominations, particularly at the higher end of the complaint scale. At lower complaint scale, the likelihood of receiving award nominations may vary depending on each officer, but as it gets to the higher complaint scale, the likelihood of receiving many award nominations seem to decrease. However, it is also important to note that there may be a pattern where officers with many complaints receive fewer number of awards, there is no clear evidence of a linearity or proportionality.

@@

```{r}
#| label: award-nomination-by-number-of-arrests
#| fig-cap: Award Nomination by Number of Arrests
#| echo: false

arr_awd <- ggplot(all_officer, aes(x = arrest, y = awd_perf)) + 
  xlab("Number of Arrests") + ylab("Award Nomination") +
  geom_point(shape=18, color="red")
